Towards understanding the central limit theorem by learning Python basics

### 2021-03-13 paper/article.tex
Update the paper according to Alex's suggestion, add references to literature

We, the authors, have been spending a reasonable time with private tutoring in many fields
of mathematics---we are teachers. And, we agree on that probability theory is maybe
the most difficult field of mathematics---if it is about \textit{understanding}. For us, rough explanation
of statistical relationships seems to be easier than explaining the delicate issues
of properties of binomial coefficients, infinite sums and convergence. Indeed, in some
sense, statistics is much easier to explain---one needs to run virtual experiments
in a web browser % add links to Steve Phelps' https://www.geogebra.org/m/bxytp4hq, Ice Cream
% https://www.geogebra.org/m/qsjbqjfe, Broken Stick
% https://www.geogebra.org/m/bu5sxbn2, Fair Coin
% + Andreas Lindner's collections: https://www.geogebra.org/m/AytaSakt, https://www.geogebra.org/m/qXPnyCTZ
and this is well supported for several years, among others, in the online version of GeoGebra.

Connecting pure mathematics and real life or virtual experiments, can be, to our experience,
extremely hard. Probability is an important item of the curriculum at secondary level,
thus it is also an important item in the teacher training. But, to our experience,
most prospective teachers \textit{never} understand the basics beyond introductory level.
Our conjecture is that the main problem lies in students' verification if theory indeed meets practice,
so, at the end of the day, a very uncertain knowledge can be observed in the students' perception. As a consequence,
in most cases, not just yesterday's and today's mathematics teachers cannot connect
theory and practice, but their students either. Roughly speaking: everybody talks about probability and
statistics, but nobody understands anything behind the scenes.

In our paper we propose a radical way, or at least, to extend the classic way how
basics of elementary probability theory should be taught. It is \textit{programming}.
Our conjecture is that understanding will be significantly improved if the main concepts
of probability theory are supported with simple computer programs. The ouput for their flexible input
should be immediately checked, eventually in a web browser, and by changing the input parameters
to a higher number, the results can be quickly generalized. We emphasize that computers
can take on the high number of computatations from human---and this is what we exactly want.
A quite simple \textit{sample space} can contain \textit{a lot} of elements, even millions or more, and it cannot
be expected that non-experts can count their elements without deeper knowledge. In fact,
experts in probability are usually experts in combinatorics as well. To reach a reasonable
level of understanding of the main concepts in probability, by using just paper and pencil,
one needs to have a very strong and \textit{safe} background in combinatorics.

@@

First author hold a series of supplementary lectures on probability theory two years ago, where second author
took part a student and illustrated several homework assignments by using Python programs.
Some of his illustrations were a bit complicated for beginners, so the idea came that
simplified versions of such programs could be helpful for many students.
But, instead of giving a problem in probability theory and asking for a solution, in this
project we reversed the logic: We revealed the Python code in advance, and wanted to know
what problem will be solved with its help.

The first example is shown in Fig.~\ref{main-py-window}. We kept the original version to highlight
the possibility to write special characters in the Python code (we used Python version 3),
including accented German letters. In the first sheet of problems each student had to try
the code on his or her own on the website \url{https://repl.it/languages/python3},
by paying attention on the correct indentation as well---the students, however, did not have
to type the code but just copy it from the email into the programming window.
We also instructed all students with an
explanatory screenshot how the program should be started and the output obtained.

@@

In fact, we created a couple of other programs in advance, but always waited for
all solutions turned in by the students before finalizing the next sheet of problems.
We \textit{did not} give any feedback on their solutions. Instead, if we felt that
some difficulties arose, then we tried to cover them in the next sheet of problems.
The reason behind this was to test if it is possible to create a set of problems
\textit{in advance} for a possible \textit{static} learning path. Even if our experiment
was not fully static, for a future project now we already have a static set of materials.

Here we emphasize that easy access to an online version of a Python interpreter
played a significant role in our project. We did not want to ask the students
to install anything on their computer. Instead, we tried to minimize the gap
between theory and practice. In fact, \url{repl.it} is not the only website
allowing the user to try his or her code freely---we were just
satisfied with its simple user interface. Later, to obtain graphical output in an easy way, we also
introduced how Sage worksheets at \url{cocalc.com} can be accessed---in this latter case,
however, accented letters did not look elegant.

@@

The students had to compare these two program codes, check if their outputs are the same
(even for 6 or 7 dice), and find the differences between the way of programming.
We actually wanted to add another connection with theory, namely, to verify if the
formula
\begin{equation}
V(X)=E((X-E(X))^2)=E(X^2)-E(X)^2
\label{variance}
\end{equation}
(also known as Steiner's \textit{Verschiebungssatz}, that is, \textit{translation theorem})
can be reflected in the two solutions.

As another activity, a third program was given (see Fig.~\ref{6c.py}) to prepare the last
topic, namely, some steps towards the central limit theorem (CLT).
\begin{figure}
\lstinputlisting[language=Python]{../progs/en_US/6c.py}
\caption{A third program in the sixth sheet (translated into English)}
\label{6c.py}
\end{figure}
Here we provide the questions to this third program:
\begin{enumerate}
\item What computation is performed by this program?
\item Consider following modifications:
\begin{lstlisting}[language=Python,firstnumber=2]
Ω = set(itertools.product({1, 2, 3, 4, 5, 6}, repeat=4))
possible_values = range(30)
\end{lstlisting}
\begin{lstlisting}[language=Python,firstnumber=8]
  heads[sum(ω)] += 1
\end{lstlisting}
Explain the output of the modified program.
\item We are about to plot the obtained data. So we change line 10 as follows:
\begin{lstlisting}[language=Python,firstnumber=10]
for w in zip(*(possible_values, probabilities)):
  print(f"{w[0]},{w[1]}")
\end{lstlisting}
Copy the output of this program in a spreadsheet software and create an XY diagram
to get some graphical output.
\item Extend the program with these two additional lines:
\begin{lstlisting}[language=Python,firstnumber=12]
distribution = [sum(probabilities[:h]) for h in possible_values]
print(list(zip(*(possible_values, distribution))))
\end{lstlisting}
Explain the result.
\end{enumerate}
In fact, the variable naming was somewhat incorrect here (because \texttt{heads} 
actually refers to the sums of the outcomes), but we wanted to make very little
changes on the code.

@@

We conjecture that making a difference between a discrete distribution and a continuous one may be
a big jump in the level of understanding for the student. Discrete distributions are \textit{never} normal,
and to speak about a Gaussian distribution one should introduce the notion of limit in a very
delicate form. Here, we should translate the random variable $X$ to have the expected value $0$ first,
and then normalize it by a function of $d$. We did not see any sign of this concept among the
students' solutions. Maybe, in a next project, we will explicitly ask the students to try to
\textit{approximate} the discrete distributions with a continuous one, by searching for the adequate
expected value and standard deviation. Also, experimenting with higher values for $d$ should also be
suggested---no student
tried to increase the value of \texttt{max\_dice}, even if technically it is possible to try
\texttt{max\_dice = 8} in a \texttt{cocalc.com} Sage session. (Fig.~\ref{10} shows the output
for \texttt{max\_dice = 10} when run natively on a recent computer, after about 1 minutes
of computation. Here we warn the reader that a sample space $\Omega$ with more than 60 millions of elements
must be observed, since $|\Omega|=6^{10}=60,466,176$.)

@@

@@

\begin{thebibliography}{99}

\end{thebibliography}

>>>>

We, the authors, have been spending a reasonable time with private tutoring in many fields
of mathematics---we are teachers. And, we agree on that probability theory is maybe
the most difficult field of mathematics---if it is about \textit{understanding}. For us, rough explanation
of statistical relationships seems to be easier than explaining the delicate issues
of properties of binomial coefficients, infinite sums and convergence. Indeed, in some
sense, statistics is much easier to explain---one needs to run virtual experiments
in a web browser
and this is well supported for several years, among others, in the online version of GeoGebra\footnote{%
See, for example, Steve Phelps' online applets
\url{https://www.geogebra.org/m/bxytp4hq},
\url{https://www.geogebra.org/m/qsjbqjfe} and
\url{https://www.geogebra.org/m/bu5sxbn2},
that visualize experiments with ice cream purchases, a broken stick or a fair coin, respectively;
or check out Andreas Lindner's collections at \url{https://www.geogebra.org/m/AytaSakt}
and \url{https://www.geogebra.org/m/qXPnyCTZ}.}.

Connecting pure mathematics and real life or virtual experiments, can be, to our experience,
extremely hard. Probability is an important item of the curriculum at secondary level,
thus it is also an important item in the teacher training. But, to our experience,
most prospective teachers \textit{never} understand the basics beyond introductory level.
Our conjecture is that the main problem lies in students' verification if theory indeed meets practice,
so, at the end of the day, a very uncertain knowledge can be observed in the students' perception. As a consequence,
in most cases, not just yesterday's and today's mathematics teachers cannot connect
theory and practice, but their students either. Roughly speaking: everybody talks about probability and
statistics, but nearly nobody understands anything behind the scenes.

In our paper we propose a radical way, or at least an extension of the classic way, how
basics of elementary probability theory should be taught. It is \textit{programming}.
Our conjecture is that understanding will be significantly improved if the main concepts
of probability theory are supported with simple computer programs. The output for their flexible input
should be immediately checked, eventually in a web browser, and by changing the input parameters
to a higher number, the results can be quickly generalized. We emphasize that computers
can take on the high number of computatations from human---and this is what we exactly want.
A quite simple \textit{sample space} can contain \textit{a lot} of elements, even millions or more, and it cannot
be expected that non-experts can count their elements without deeper knowledge. In fact,
experts in probability are usually experts in combinatorics as well. To reach a reasonable
level of understanding of the main concepts in probability, by using just paper and pencil,
one needs to have a very strong and \textit{safe} background in combinatorics.

@@

First author hold a series of supplementary lectures on probability theory two years ago, where second author
took part as a student and illustrated several homework assignments by using Python programs.
Some of his illustrations were a bit complicated for beginners, so the idea came that
simplified versions of such programs could be helpful for many students.
But, instead of giving a problem in probability theory and asking for a solution, in this
project we reversed the logic: We revealed the Python code in advance, and wanted to know
what problem will be solved with its help.

\subsection{Why Python?}

Why did we choose Python as the programming language for this course?
There are a couple of reasons we would like to note. First of all: It is
free, independent of the used operating system, has good documentation
and great tutorials. In addition, Python is currently a very popular
programming language: according to
\url{https://madnight.github.io/githut/#/pull_requests/2020/4}, Python
was the second most popular language on GitHub in the 4th quarter of 2020;
moreover, ZDNet's recent article \cite{LiamTung2020} highlights that Python
was the second best popular language in November 2020.
Many programming courses teach it as a first
programming language because of its ease of use and short expressive
code.

Also, we think that Python is a great tool to express mathematical
concepts in a concise and direct way---several recent books give
examples on a possible start, see e.g.~\cite{Fuhrer2016}, \cite{Mehta2015} and \cite{Stewart2014}.
In fact, in our course students
needed to learn classic programming concepts like conditions and loops
together with probability, so we needed a programming language with a
steep learning curve. We think Python can be seen as a great tool that
supports students with the understanding of mathematical probability
concepts without the need to learn too much programming.

On the other hand, in the field of data science
Python is currently the mostly used programming language. Especially in
machine learning, which could be seen as an advanced application of
probability, Python is very popular. We think that students who believe
that the exercises they solve are not artificially constructed for the
course, but have some connection to the real world, will be more
motivated. As a teacher we can give students the impression that if they
use Python in their introductory course on probability, they take a first
step into the prestigious field of data science. This may increase the
motivation of students and as a result they will achieve higher.

\subsection{The programs}

The first example of the introduced Python programs
is shown in Fig.~\ref{main-py-window}. We kept the original version to highlight
the possibility to write special characters in the Python code (we used Python version 3),
including accented German letters. In the first sheet of problems each student had to try
the code on his or her own on the website \url{https://repl.it/languages/python3},
by paying attention on the correct indentation as well---the students, however, did not have
to type the code but just copy it from the email into the programming window.
We also instructed all students with an
explanatory screenshot how the program should be started and the output can be obtained.

@@

In fact, we created a couple of other programs in advance, but always waited for
all solutions to be turned in by the students before finalizing the next sheet of problems.
We \textit{did not} give any feedback on their solutions. Instead, if we felt that
some difficulties arose, then we tried to cover them in the next sheet of problems.
The reason behind this was to test if it is possible to create a set of problems
\textit{in advance} for a possible \textit{static} learning path. Even if our experiment
was not fully static, for a future project now we already have a static set of materials.

Here we emphasize that easy access to an online version of a Python interpreter
played a significant role in our project. We did not want to ask the students
to install anything on their computers. Instead, we tried to minimize the gap
between theory and practice. In fact, \url{repl.it} is not the only website
allowing the user to try his or her code freely---we were just
satisfied with its simple user interface. Later, to obtain graphical output in an easy way, we also
introduced how Sage worksheets at \url{cocalc.com} can be accessed---in this latter case,
however, accented letters did not look elegant.

@@

The students had to compare these two program codes, check if their outputs are the same
(even for 6 or 7 dice), and find the differences between the way of programming.
We actually wanted to add another connection with theory, namely, to verify if the
formula
\begin{equation}
V(X)=E((X-E(X))^2)=E(X^2)-E(X)^2
\label{variance}
\end{equation}
(also known as Steiner translation theorem in Germany and König-Huygens formula in the French literature)
can be reflected in the two solutions.

As another activity, a third program was given (see Fig.~\ref{6c.py}) to prepare the last
topic, namely, some steps towards the central limit theorem (CLT, \cite{Polya1920}).
\begin{figure}
\lstinputlisting[language=Python]{../progs/en_US/6c.py}
\caption{A third program in the sixth sheet (translated into English)}
\label{6c.py}
\end{figure}
Here we provide the questions to this third program:
\begin{enumerate}
\item What computation is performed by this program?
\item Consider following modifications:
\begin{lstlisting}[language=Python,firstnumber=2]
Ω = set(itertools.product({1, 2, 3, 4, 5, 6}, repeat=4))
possible_values = range(30)
\end{lstlisting}
\begin{lstlisting}[language=Python,firstnumber=8]
  heads[sum(ω)] += 1
\end{lstlisting}
Explain the output of the modified program.
\item We are about to plot the obtained data. So we change line 10 as follows:
\begin{lstlisting}[language=Python,firstnumber=10]
for w in zip(*(possible_values, probabilities)):
  print(f"{w[0]},{w[1]}")
\end{lstlisting}
Copy the output of this program in a spreadsheet software and create an XY diagram
to get some graphical output.
\item Extend the program with these two additional lines:
\begin{lstlisting}[language=Python,firstnumber=12]
distribution = [sum(probabilities[:h]) for h in possible_values]
print(list(zip(*(possible_values, distribution))))
\end{lstlisting}
Explain the result.
\end{enumerate}
In fact, the variable naming was somewhat incorrect here (because \texttt{heads} 
actually refers to the sums of the outcomes), but we wanted to make very little
changes on the code.

@@

At this point we need to give some description on which form of CLT should be expected to discover by the students.
In fact, CLT has several versions: both formal and informal variants. Roughly speaking, CLT can be summarized
in the following form (see \cite{Wikipedia-CLT}):
\begin{quote}
When independent random variables are added, their properly normalized
sum tends toward a normal distribution (informally a bell curve) even if
the original variables themselves are not normally distributed.
\end{quote}
A formal definition given in \cite[p.~241]{Montgomery2014} is as follows:
\begin{quote}
If $X_{1},X_{2},\ldots,X_{n}$
are $n$ random samples drawn from a population
with overall mean $\mu$ and finite variance $\sigma^2$,
and if ${\bar {X}}_{n}$ is the sample mean, the
limiting form of the distribution, $$Z=\lim _{n\to \infty}{\sqrt {n}}{\left({\frac {{\bar {X}}_{n}-\mu }
{\sigma}}\right)},$$ is the standard normal
distribution.
\end{quote}
In fact, we avoided going into any formal detail during our course by assuming that the provided formula
here \textit{may not promote understanding}.

More importantly, we conjecture that making a difference between a discrete distribution and a continuous one may be
a big jump in the level of understanding for the student. Discrete distributions are \textit{never} normal,
and to speak about a Gaussian distribution one should introduce the notion of limit in a very
delicate form. Here, we should translate the random variable $X$ to have the expected value $0$ first,
and then normalize it by a function of $d$. We did not see any sign of this concept among the
students' solutions. Maybe, in a next project, we will explicitly ask the students to try to
\textit{approximate} the discrete distributions with a continuous one, by searching for the adequate
expected value and standard deviation. Also, experimenting with higher values for $d$ should also be
suggested---no student
tried to increase the value of \texttt{max\_dice}, even if technically it is possible to try
\texttt{max\_dice = 8} in a \texttt{cocalc.com} Sage session. (Fig.~\ref{10} shows the output
for \texttt{max\_dice = 10} when run natively on a recent computer, after about 1 minutes
of computation. Here we warn the reader that a sample space $\Omega$ with more than 60 millions of elements
must be observed, since $|\Omega|=6^{10}=60,466,176$.)

@@

We emphasize that this research just the first step in a longer research project.
Basically, we would like to continue with interviews with the participating
students whether their knowledge and motivation were improved during the course.
This will give us further feedback to meet the best decision towards further
refinements of the course.

@@

\bibliographystyle{tfs}
\bibliography{article}
